---
title: "Supplementary Materials to Chapter 14: Persistence of Passive Immunity, Natural Immunity (and Vaccination)"
author: "Amy K. Winter & C. Jessica E. Metcalf"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary
In this supplement to Chapter 14 titled 'Persistence of Passive Immunity, Natural Immunity (and Vaccination),' we illustrate some examples of fitting models to serological data including i) fitting waning of maternal immunity (with simulated data, generated here); ii)  fitting the age profile of seropositivity, across a range of age class designs (with simulated data, available as 'simulated.seroprevalence.profile.csv'); iii) extracting the force of infection (again with 'simulated.seroprevalence.profile.csv'); and iv) exploring a case study for rubella (using existing serology data extracted from the literature, available as 'serology.data.csv' combined with data from United Nations Population Division, 2015 on fertility over age 'asfr.data.csv' and population structure 'female.pop.data.csv').

## Methods for analysis of IgG titers
Following Waaijenborg et al. (2013), we can define a function that captures decay of maternal antibody titers with time (ignoring for simplicity fluctuations in status associated with vaccinated individuals):

```{r maternalImmunity1}
waning <- function(age, m, d, b){
  log(m*exp(-d*age)+b)
}
```

We can generate data according to this function assuming that log titers are normally distributed (here parameters reflect those estimated for measles in Waaijenborg et al. (2013)):

```{r maternalImmunity2}
ages <- seq(1,60,length=100)/12 
dist.titers <- rnorm(length(ages),waning(ages,m=1.61,d=7.77,b=0.011),sd=1.11)
```
and then write down a likelihood function for this data (returning the negative log likelihood, as minimization is more tractable than maximization): 

```{r maternalImmunity3}

like.waning <- function(par,age,titer){
  par<- exp(par)  #simple way of ensuring all parameters are positive
  pred <- waning(age=age, m=par[1], d=par[2], b=par[3])
  u <- dnorm(titer, pred,par[4],log=TRUE)
  return(-sum(u))
}

```

We can then try and recover these parameters by applying optimization techniques to this data with the function 'optim': 

```{r maternalImmunity4}
tmp <- optim(par=log(c(1,1,0.1,1)),like.waning,age=ages,titer=dist.titers, 
             method="Nelder-Mead")
 #check fit has converged - should be 0
print(tmp$convergence)
#compare to the values set in the simulation: 
print(cbind(estimated=round(exp(tmp$par),2),true=c(1.61,7.77,0.011,1.11))) 
```

We can plot out these results, showing the data (filled points) and the 'true' relationship (grey curve); and the fitted relationship (red curve). This indicates that even if parameters recovered are not precisely those used in the simulation (see above), the curves are  similar (and further optimization might also move the parameters closer to the 'true' values): 

```{r maternalImmunity5}
plot(ages, dist.titers, xlab="Ages (years)", ylab="Titers", pch=19)
points(ages,waning(ages,m=1.61,d=7.77,b=0.011), type="l",col="grey")
points(ages,waning(ages,m=exp(tmp$par[1]),d=exp(tmp$par[2]),b=exp(tmp$par[3])), 
       type="l",col="red")
```

## Methods for analysis of seropositivity 

We move from analysis of antibody titers (requiring techniques associated with analysis of continuous variables) to considering analysis of seropositivity, proportion of individuals with titers above a defined threshold (generally established as being reflective of protection from infection), usually within an age class. This requires techniques for analysis of binary variables. 

The design of serological surveys oftens reflects a choice of age bins (generally dictated by logistical constraints), and the implications of this choice are not always fully considered. In particular, grouping younger ages when the shape of the seropositivity curve is the steepest can result in underestimates of seropositivity across age. We illustrate this using simulated data, based around a model introduced in Metcalf et al. 2012, assuming age contact based around POLYMOD estimates from Mossong et al. 2008, with $R_0$=6, and a crude birth rate set to 30 per 1000 people. For each age (corresponding to individuals at the exact age in years), the simulation provides the number out of 100 individuals that are seropositive, and the number that are seronegative. We use this to estimate the mean proportion that would have been seropositive if data from a serological survey had been available in 5 year and 10 year age bins (corresponding to Figure 2A in Chapter 15). 

```{r Estimating Age-Specific seropositivity with Simulated Data}
#Figure 2A Code
sim.data <- read.csv("./data/simulated.seroprevalence.profile.csv")
prop.seropos <- sim.data$Pos/sim.data$Tot

index <- findInterval(1:49,seq(5,45,5))+1
prop.seropos.5year <- rep(NA,10)
for (a in 1:10){
  prop.seropos.5year[a] <- mean(prop.seropos[index==a])
}

index <- findInterval(1:49,seq(10,50,10))+1
prop.seropos.10year <- rep(NA,5)
for (a in 1:5){
  prop.seropos.10year[a] <- mean(prop.seropos[index==a])
}

mid.age.choices <- list(seq(1,49,1), c(2.5,seq(7,47,5)), c(5,seq(14.5,44.5,10))) 
```

To explore the implications of grouping age classes, we plot the proportion seropositive by age, and illustrate the overall pattern by fitting a line using a smoothed spline. As age bins widen across the ages where the seroprevalence curve is concave (ages 5-15 years), our underestimation of the proportion seropositive increases.


```{r plot Figure 2A, fig.width=6, fig.height=6}
par(mar=c(5,4,4,5)+.1, bty="l",pty="s")
plot(mid.age.choices[[1]], prop.seropos, type="p", pch=16, col="black", xlim=c(0,15), 
     xlab="Age (yrs)", ylab="Proportion seropositive", cex.lab=1.5, cex.axis=1.5)
lines(1:15, predict(smooth.spline(mid.age.choices[[1]][1:15], 
                                  prop.seropos[1:15]), 1:15)$y, col="black", lty=1)
points(mid.age.choices[[2]], prop.seropos.5year[1:10], col="dark grey", pch=17, cex=1.2)
lines(2.5:15, predict(smooth.spline(mid.age.choices[[2]][1:7], 
                                    prop.seropos.5year[1:7]), 2.5:15)$y, col="dark grey", 
      lty=2)
points(mid.age.choices[[3]], prop.seropos.10year[1:5], col="black", pch=18, cex=1.2)
lines(5:15, predict(smooth.spline(mid.age.choices[[3]][1:4], 
                                  prop.seropos.10year[1:4]), 5:15)$y, col="black", lty=3)
legend("topleft", c("annual age bins","five year age bins", "ten year age bins"),
       lty=c(1,2,3), cex=1.1, col=c("black","dark grey", "black"), pch=c(16,17,18))
```

The power of seropositivity data is the insights it can yield into the underlying dynamics. Returning to the highest resolution data (1 year age bins), we can fit models to the age profiles of seropositivity, and use these to infer the pattern of the force of infection over age. Here, we illustrate  different parametric methods: fractional polynomials,  piecewise constant, semi-parametric penalized regression splines, and parametric local polynomials. This code is based on Hens et al. (2012), framed below in the form of functions that can be applied to data in the form provided. 

```{r Estimating Age-Specific Seroprevalence and ASFOI with Simulated Data}

#Figure 2B Code
### Function to Extract ASFOIs for a serosurvey using fractional polynomials
###
##' @param - pos - number of positives for each age
##' @param - Tot - number of samples for each age
##' @param - Age - age in years
###
### outputs estimated ASFOIs and Fitted Seroprevalence 
ExtractASFOIEstimates.FP <- function(pos, Tot, Age){
  
  #determing to use first or second degree
  best.1 <- search.fracpoly.one(y=pos,tot=Tot,x=Age)
  best.2 <- search.fracpoly.two(y=pos,tot=Tot,x=Age)
  firstdegree <- (best.1$deviance - best.2$deviance < 4.605) 
  print(c(firstdegree, "Use First Degree:"))
  #chi square with 2 degrees of freedom 90% CI, 
  if (firstdegree){
    if (best.1$power!=0){
      fit.best <- glm(cbind(pos,Tot-pos) ~ I(Age^best.1$power), 
                      family=binomial(link="logit")); print("power!=0")
    } else {
      fit.best <- glm(cbind(pos,Tot-pos) ~ I(log(Age)), 
                      family=binomial(link="logit")); print("power=0")
    }
  } else {
    if (best.2$power[1]==best.2$power[2]){
      fit.best <- glm(cbind(pos,Tot-pos) ~ I(Age^(best.2$power[1])) + 
                        I(Age^(best.2$power[2])*log(Age)), family=binomial(link="logit")) 
    } else {
      if (best.2$power[1]==0){
        fit.best <- glm(cbind(pos,Tot-pos) ~ I(log(Age)) + 
                          I(Age^(best.2$power[2])), family=binomial(link="logit")) 
      } else if (best.2$power[2]==0){
        fit.best <- glm(cbind(pos,Tot-pos) ~ I(Age^(best.2$power[1])) + 
                          I(log(Age)), family=binomial(link="logit")) 
      } else {
        fit.best <- glm(cbind(pos,Tot-pos) ~ I(Age^(best.2$power[1])) + 
                          I(Age^(best.2$power[2])), family=binomial(link="logit")) 
      }
    }
  }
  
  age.grid=seq(1,max(Age),by=1)
  p=predict(fit.best, newdata=data.frame(Age=age.grid),type="response", se.fit=T)
  fitted <- p$fit
  asfoi <-foi.num(x=age.grid,p=fitted)
  
  return=list(asfoi=asfoi$foi, 
              ages=asfoi$grid, 
              fitted=fitted)
  
}

### Function to search for the best 1st degree polynomial
###
##' @param - y - the number of seropos per age, x
##' @param - tot - the number of sample per age, x
##' @param - x - age
##' @param - mc - monotonicity constraint, assumes x is ordered
###
### outputs the "best" power (lowest deviance), its associated deviance
### and the number of non-converged models for the first degree polynomial
search.fracpoly.one<-function(y,tot,x,mc=TRUE){
  pow1<-seq(-2,3,0.01) #searching through these powers
  deviance<-deviance(glm(cbind(y,tot-y)~x, family="binomial"(link=logit)))
  #we compare all models to the baseline deviance of the null model
  power<-1 #only looking at 1st degree polynomials
  mistake<-NULL
  
  for (i in 1: (length(pow1))){
    if(pow1[i]==0){term1<-log(x)} else{term1<-(x)^(pow1[i])} 
    glm.try<- suppressWarnings(glm(cbind(y,tot-y)~term1, family="binomial"(link=logit)))
    if(glm.try$converged==FALSE){mistake<-rbind(mistake, c(1,pow1[i]))}
    else{#replace deviance if this one is better -> ordered approach to finding best model
      if(deviance(glm.try)<deviance){ 
        if (((mc)&&(sum(diff(predict(glm.try))<0)==0))|(!mc)){
          deviance<-deviance(glm.try)
          power<-pow1[i]
        }
      }
    }
  }
  return(list(power=power, deviance=deviance, mistake=mistake))
}


### Function to search for the best 2nd degree polynomial
###
##' @param - y - the number of seropos per age, x
##' @param - tot - the number of sample per age, x
##' @param - x - age
##' @param - mc - monotonicity constraint, assumes x is ordered
###
### outputs the "best" powers (lowest deviance), its associated deviance
### and the number of non-converged models for the second degree polynomial
search.fracpoly.two<-function(y,tot,x,mc=TRUE){
  pow<-seq(-2,3,0.1)
  deviance<-deviance(glm(cbind(y,tot-y)~x+I(x^2), family="binomial"(link=logit)))
  mistake<-NULL
  
  for (i in 1: (length(pow))){
    for (j in i: (length(pow))){
      if(pow[i]==0){term1<-log(x)} else{term1<-(x)^(pow[i])}
      if(pow[j]==pow[i]){term2<-term1*log(x)} 
      else if(pow[j]==0){term2<-log(x)} 
      else{term2<-(x)^(pow[j])}
      glm.try<-glm(cbind(y,tot-y)~term1+term2, family="binomial"(link=logit))
      if(glm.try$converged==FALSE){mistake<-rbind(mistake, c(1,pow[i],pow[j]))}
      else{
        if(deviance(glm.try)<deviance){
          if (((mc)&&(sum(diff(predict(glm.try))<0)==0))|(!mc)){
            deviance<-deviance(glm.try)
            power<-c(pow[i],pow[j])
          } 
        }
      }
      
    }
  }
  return(list(power=power, deviance=deviance, mistake=mistake))
}


### Function to Extract ASFOIs for all serosurveys using Semi-parametric 
### penalized smoothing splines
###
##' @param - pos - number of positives for each age
##' @param - Tot - number of samples for each age
##' @param - Age - age in years
###
### outputs estimated ASFOIs and Fitted Seroprevalence
ExtractASFOIEstimates.GAM <- function(pos, Tot, Age){
  
  #Smooth then constrain approach
  
  #ungroup the data
  df <- rbind(data.frame(a=rep(Age, c(pos)), y=1), 
              data.frame(a=rep(Age, c(Tot-pos)), y=0))
  y<-df$y[order(df$a)]
  a<-df$a[order(df$a)]
  
  fit.best <-gam(y~s(a))
  #gam.check(fit.best)
  
  age.grid=seq(1,max(Age),by=1)
  p=predict(fit.best,newdata=data.frame(a=age.grid),type="response", se.fit=T)
  fitted <- p$fit[1:49]
  asfoi.orig <-foi.num(x=age.grid,p=fitted)
  
  #issue with monotonicity dealt with up front using the 'Pool Adjacent Violator Algorithm"
  #data is pooled if not always increasing, otherwise data remains the same
  Monotonicity.pos <- pavit(pos=round(fitted*Tot),tot=Tot)$pai2
  asfoi <-foi.num(x=age.grid,p=Monotonicity.pos)
  
  
  return=list(asfoi=asfoi$foi, 
              ages=asfoi$grid, 
              fitted=fitted,
              Monotonicity.pos=Monotonicity.pos,
              asfoi.orig=asfoi.orig$foi)
  
}


### Function pavit - Extracted directly from Hens et al. 2012 textbook
### The pool adjacent violator algorithm in R
### data represents the ordered fitted values
pavit<- function(pos=pos,tot=rep(1,length(pos)))
{
  gi<- pos/tot
  pai1 <- pai2 <- gi
  N <- length(pai1)
  ni<-tot
  for(i in 1:(N - 1)) {
    if(pai2[i] > pai2[i + 1]) {
      pool <- (ni[i]*pai1[i] + ni[i+1]*pai1[i + 1])/(ni[i]+ni[i+1])
      pai2[i:(i + 1)] <- pool
      k <- i + 1
      for(j in (k - 1):1) {
        if(pai2[j] > pai2[k]) {
          pool.2 <- sum(ni[j:k]*pai1[j:k])/(sum(ni[j:k]))
          pai2[j:k] <- pool.2
        }
      }
    }
  }
  return(list(pai1=pai1,pai2=pai2))
}

### Function to get ASFOI from fitted values of seroprevalence by age 
### Taken directly from Hens et al. 2012 textbook
###
##' @param - x - age
##' @param - p - fitted seroprevalence associated with age, x in order
###
### outputs ASFOI
foi.num<-function(x,p)
{
  grid<-sort(unique(x))
  pgrid<-(p[order(x)])[duplicated(sort(x))==F] 
  dp<-diff(pgrid)/diff(grid) #slope
  numerator <-approx((grid[-1]+grid[-length(grid)])/2,dp,grid[c(-1,-length(grid))])$y
  denominator <- (1-pgrid[c(-1,-length(grid))])
  foi <- numerator/denominator
  return(list(grid=grid[c(-1,-length(grid))],foi=foi,
              derivative=numerator,
              prop.seroneg=denominator))
}

### Function to Extract ASFOIs for all serosurveys
###
##' @param - pos - number of positives for each age
##' @param - Tot - number of samples for each age
##' @param - Age - age in years
##' @param - age.break - age breaks for the piecewise function
###
### outputs estimated ASFOIs and Fitted Seroprevalence
ExtractASFOIEstimates.piecewise.agebreak <- function(pos, Tot, 
                                                     Age, age.break=c(0,seq(5,50,5))){
  
  df <- data.frame(cbind(age=Age,Tot=Tot, Neg=Tot-pos, Pos=pos))
  
  #number of time optimizing starting at random values, 
  ##   and then starting from mean dist from random values
  noptims <- 20
  asfoi.bytest <- array(NA, dim=c((length(age.break)-1), noptims))
  for (test in 1:noptims){
    set.seed(test)
    pars <- abs(rnorm((length(age.break)-1), 0.11, 0.05))
    out <- optim(log(pars), fn=ll.data.as.foi, data=df, age.break=age.break)
    asfoi.bytest[,test] <- exp(out$par)
  }
  asfoi.mean <- apply(asfoi.bytest, 1, FUN=mean, na.rm=T)
  out <- optim(log(asfoi.mean), fn=ll.data.as.foi, data=df, age.break=age.break)
  asfoi.piecewise <- exp(out$par)
  
  #turn piecewise ASFOI into ASFOI over all ages
  ages.asfoi <- 1:49
  index <- findInterval(ages.asfoi-1, age.break)
  asfoi <- asfoi.piecewise[index]
  
  #get predicted seroprevalence predicted values
  fitted <- 1-(exp(-cumsum(asfoi)))
  
  return=list(asfoi=asfoi,
              ages=ages.asfoi,
              fitted=fitted)
  
}


### -LogLikelihood by which to estimate the age specific force of infection (ASFOI)
### using piecewise constant hazard without numerical integration
### must start after maternal immunity - assume everyone is susceptible by age 1yo
###
##' @param - log.lambdas - starting parameter for age specific FOI
##' @param - data - dataframe with age, positives, and number samples per age
##' @param - age.break - upper ages used in piecewise fitting
###
### returns negative log likelihood
ll.data.as.foi  <-  function(log.lambdas, data, age.break, mort=c()){
  
  par <- log.lambdas
  dur <- c(diff(age.break)) 
  cate <- age.break
  cate <- cate[-length(cate)] 
  
  ll <- 0
  for(a in 1:length(data$age)){
    dummy2 <- data$age[a]>cate & !c(data$age[a]>cate[-1], FALSE) 
    dummy1 <- c(data$age[a]>cate, FALSE)[-1] 

    inte <- sum(dur*exp(par)*dummy1)+(exp(par[dummy2])*(data$age[a]-cate[dummy2]))
    
    p <- pmax(pmin(1-exp(-inte),1-1e-10),1e-10)
    # likelihood
    ll <- ll+dbinom(data$Pos[a],data$Tot[a],p,log=TRUE)
  }
  return(-ll)
}
```

Using the functions developed above, we can estimate the pattern of age-seropositivity, and from this the force of infection over age. We organize the data and proceed accordingly:  

```{r Running Figure 2B}
#Simulated data
npos <- sim.data$Pos
Nsamp <- sim.data$Tot
nneg <- sim.data$Neg
ages <- sim.data$age

###local polynomial fit - non-parametric
require(locfit)
lpfit <- locfit(npos/Nsamp~lp(ages,nn=0.7,deg=2),family="binomial", subset=ages>=1)
lpfitd1 <- locfit(npos/Nsamp~lp(ages,nn=0.7,deg=2),deriv=1,family="binomial",
                  alpha=0.7,deg=2, subset=ages>=1)
lpfoi <- (predict(lpfitd1,newdata=data.frame(ages=1:49)))*
         (predict(lpfit,newdata=data.frame(ages=1:49)))
lp.fitted <- (predict(lpfit,newdata=data.frame(ages=1:49)))

###fractional polynomials - parametric
fp <- ExtractASFOIEstimates.FP(pos=npos, Tot=Nsamp, Age=ages)

###penalized regression splines - semi-parametric
library(mgcv) 
gam <- ExtractASFOIEstimates.GAM(pos=npos, Tot=Nsamp, Age=ages)

###piecewise constant - flexible parametric 
#the following line of code can take up to 3 minutes to optimize
pw <- ExtractASFOIEstimates.piecewise.agebreak(pos=npos, Tot=Nsamp, 
                                               Age=ages, age.break=c(0,seq(5,50,5)))
```

To compare the inference from the different approaches, we plot them all on the same figure: 

```{r Plotting Figure 2B, fig.width=6, fig.height=6}
par(mar=c(5,4,4,5)+.1, bty="l",pty="s")
plot(ages, npos/Nsamp, pch=19, xlab="Age (yrs)", ylab="Proportion seropositive", 
     type="n", xlim=c(0,30), ylim=c(0,1), cex.lab=1.5, cex.axis=1.5)
points(ages, npos/Nsamp, type="p", cex=1.7, col="grey", pch=19)
axis(4, c(0,0.1,0.2,0.3), c(0,0.1,0.2,0.3), cex.axis=1)
mtext("Force of Infection", side=4, line=3, at=0.15, cex=1.25)
lines(1:49,lp.fitted, lwd=2, lty=1)
lines(1:49, pmax(lpfoi,0), lwd=2, col=2)
lines(1:49, fp$fitted, lwd=2, lty=2)
lines(fp$ages, fp$asfoi, lwd=2, lty=2, col=2)
lines(1:49, gam$Monotonicity.pos, lwd=2, lty=3)
lines(gam$ages, gam$asfoi, lwd=2, lty=3, col=2)
lines(1:49, pw$fitted, lwd=2, lty=4)
lines(1:49, pw$asfoi, lwd=2, lty=4, col=2)
legend("right", c("fractional polynomials","penalized regression splines", 
                  "local polynomials", "piecewise constant"),
       lty=c(2,3,1,4), cex=1)
```


## Rubella case study

We illustrate the use of serological surveys to address public health issues around rubella, with three different examples from the literature (Ouattara et al. 1986; Modarres et al. 1996 and Nessa et al. 2008); each paired with data from the United Nations Population Division (2015) yielding the age specific fertility rate, and the age distribution of the population for that geographic region at that time. Both of these are necessary to estimate the burden of Congenital Rubella Syndrome. We start by bringing in this data: 

```{r Rubella Case Study}
data <- read.csv("./data/serology.data.csv")
#WCB = women of childbearing age
WCB.pop <- read.csv("./data/female.pop.data.csv", header=TRUE, 
                    row.names=1) 
#ASFR = age specific fertility rate per 1 person
ASFR <- read.csv("./data/asfr.data.csv", header=TRUE, row.names=1) 
country <- colnames(WCB.pop)
```

Using the functions defined above, we can loop through each of these countries, fit the age profile of seropositivity using local polynomial estimators based on the function defined above (chosen as an example - other functional forms are possible (Hens et al. 2012)), and from this, infer the force of infection over age. We can plot each of these and additionally estimate the burden of CRS using methods defined in the Chapter 15. 

```{r Plotting Figure 3, fig.width=4, fig.height=4}
library("locfit")
for (j in 1:3) {
  
  study <- data$study
  
  Nsamp <- data[data$study==j,"N"]
  N <- sum(Nsamp)
  npos <- round(Nsamp*data[data$study==j,"seropositive.immune"])
  nneg <- Nsamp-npos
  mid.ages <- apply(data[study==j,7:8], 1, mean) #mid-age of age range
  max.mid.ages <- mid.ages[length(mid.ages)]
  
  #plot out data
  plot(1, 1, pch=19, xlab="Age (yrs)", ylab="Proportion seropositive", 
       type="n", xlim=c(0,55), ylim=c(0,1))
  points(mid.ages, data[study==j,5], type="p", cex=log(data[study==j,4]/10), col="grey", 
         pch=19)
  jj <- which(study==j)[1]
  title(paste(as.character(data[jj,6]), ", ",as.character(data[jj,3]), sep=""))
  mtext(paste("(N=",N,")",sep=""), 3, cex=0.7)
  axis(4, c(0,0.05,0.1,0.15,0.2, 0.25), c(0,0.05,0.1,0.15,0.2, 0.25))
  mtext("Force of Infection", side=4, line=3, at=0.1, cex=0.7)
  
  ## estimate age-specific force of infection (FOI) and CRS incidence for each country
  #local polynomial fit
  lpfit <- locfit(npos/Nsamp~lp(mid.ages,nn=0.7,deg=2),
                  family="binomial", subset=mid.ages>=1)
  lpfitd1 <- locfit(npos/Nsamp~lp(mid.ages,nn=0.7,deg=2),deriv=1,
                    family="binomial",alpha=0.7,deg=2, subset=mid.ages>=1)
  lpfoi <- (predict(lpfitd1,newdata=data.frame(mid.ages=1:max.mid.ages)))*
           (predict(lpfit,newdata=data.frame(mid.ages=1:max.mid.ages)))
  
  ## add to plot
  points(1:max.mid.ages,(predict(lpfit,newdata=data.frame(mid.ages=1:max.mid.ages))), 
         type="l",lwd=1, lty=1)
  points(1:max.mid.ages,pmax(lpfoi,0), type="l", col="darkgrey", lty=2)
  
  ## estimate CRS incidence rate
  prop.seropos.WCB <- predict(lpfit,newdata=data.frame(mid.ages=15:49))
  foi.WCB <- c(lpfoi[15:length(lpfoi)], rep(lpfoi[length(lpfoi)], 
                                            max(0,(49-length(lpfoi)))))[1:35]
  #CRS incidence by WCB age
  I.crs.byage <- (1-prop.seropos.WCB)*(1-exp(-foi.WCB*16/52))*0.65*100000
  #births per maternal age group (15-19, 20-24, 25-29, 30-34, 35-39, 40-44, 45-49)
  births.WCB.byage <- rep(WCB.pop[,j]*ASFR[,j],each=5)/5
  #CRS incidence 
  I.crs <- sum(I.crs.byage*births.WCB.byage)/sum(births.WCB.byage)
  print(paste(country[j], ", CRS incidence: ", round(I.crs, 2), " per 100,000", sep=""))
  
  ## estimate average age of infection, excluding ages >30
  age.test.1 <- seq(1,30,length=5000)
  freq <- 1-predict(lpfit,newdata=data.frame(mid.ages=age.test.1))
  freq <- freq/sum(freq)   #rescale frequency
  avg.age <- sum(freq*(age.test.1+diff(age.test.1)[1]))
  
  #get R0 using simple (G/A)
  ## Crude birth rate per 1000 from the World Bank for each country and year: 
  ## Ivory Coast 1987, Iran 1996, Bangladesh 2008
  birth.rate <- c(44.483,22.549,22.262)
  R0 <- (1000/birth.rate[j])/avg.age
  print(paste("R0", round(R0), "; Average age", round(avg.age,1),sep=" "))

}
```

## Conclusion
In this supplement to Chapter 14, we provide some basic methods for analysis of serological data, focusing on directly transmitted immunizing infections such as measles and rubella. For extensive details on these methods, and a number of important extensions, see Hens et al. 2012. 

## References

Hens, N., Z. Shkedy, M. Aerts, C. Faes, P.V. Damme, and P. Beutels, Modeling infectious disease parameters based on serological and social contact data : a modern statistical perspective. . 2012, New York: Springer.

Metcalf CJE, Lessler J, Klepac P, Morice A, Grenfell BT, Bjornstad ON. Structured models of infectious disease: Inference with discrete data. Theor Popul Biol 2012. 82: p. 275-282.

Modarres, S., and N.N. Oskoii, Immunity of children and adult females to rubella virus infection in Tehran. Iran. j. méd. sci, 1996. p. 69-73.

Mossong, J., N. Hens, M. Jit, P. Beutels, K. Aranen, R. Mikolajczyk, M. Massari, S. Salmaso, G. Scalia Tomba, J. Wallinga, J. Heijne, M. Sadkowska-Todys, M. Rosinska, and W.J. Edmunds, Social Contacts and Mixing Patterns Relevant to the Spread of Infectious Diseases. PloS Medicine, 2008. 5: p. e74.

Nessa, A., M. Islam, S. Tabassum, S. Munshi, M. Ahmed, and R. Karim, Seroprevalence of rubella among urban and rural Bangladeshi women emphasises the need for rubella vaccination of pre-pubertal girls. Indian journal of medical microbiology, 2008. 26(1): p. 94.

Ouattara, S., J. Brettes, R. Kodjo, K. Penali, G. Gershy-Damet, A. Sangare, Y. Aron, and V. Akran, Seroepidemiology of rubella in the Ivory coast. Geographic distribution. Bulletin de la Societe de pathologie exotique et de ses filiales, 1986. 80(4): p. 655-664.

Population Division, Department of Economic, Social Affairs and United Nations. (2015). wpp2015: World Population Prospects 2015. R package version 1.0-1. http://CRAN.R-project.org/package=wpp2015

Waaijenborg, S., S.J. Hahné, L. Mollema, G.P. Smits, G.A. Berbers, F.R. van der Klis, H.E. de Melker, and J. Wallinga, Waning of maternal antibodies against measles, mumps, rubella, and varicella in communities with contrasting vaccination coverage. Journal of Infectious Diseases, 2013. 208(1): p. 10-16


